---
title: "trip_validation_melb"
author: "Marina_Qin_Steve_Carl"
date: "2025-##-##"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages(c('tidyverse','data.table','gt')
library(tidyverse)
library(ggplot2)
library(dplyr)
library(data.table)
library(gt)

rm(list=ls())
```

## Load baseline trips.csv and VISTA trips
```{r load-jibe-trips, message=FALSE}
## Select model outputs to validate (e.g. Melbourne model outputs of trips.csv and pp_2021.csv, respectively)
trips_ref <- file.choose() %>% fread() 
```

```{r load-jibe-persons, message=FALSE}
pp_ref <- file.choose() %>% fread()
```

```{r load-vista-trips, message=FALSE}
## Select reference survey data (e.g. VISTA trips and persons data; T_VISTA_1220_LGA_V1.csv and P_VISTA_1220_LGA_V1.csv
trips_vista <- file.choose() %>% fread() 
```

```{r load-vista-persons, message=FALSE}
pp_vista <- file.choose() %>% fread() 
```

```{r load-vista-households, message=FALSE}
hh_vista <- file.choose() %>% fread() 
```

## Data Preparation
Note that 
1. For baseline data, home-based trips and RRT are factored/weighted by 2

2. Since VISTA is a one-day diary, all trips are factored/weighted by 7, to match the JIBE model full-week modelling approach. 

> - *SP: to confirm that the Melbourne model output will cover a full week like the Manchester model output*
> - *CH: this is my understanding.  Another thought is, instead of treating the one day as 'any day', should we instead try to stratify validation for specific days of the week, at least for weekend vs weekday.  We could do this using household information:*
> |Variable|Type|Description|
> |---|---|---|
> |surveyperiod|char|Year the household was surveyed|
> |travdow|char|Household travel day - day-of-week|
> |travmonth|int|Month of survey|
> |daytype|char|Survey day sorted into weekday or weekend day types|
> - a follow up thought is, why don't we use the day stratified trip information to inform distribution estimates day by day, instead of using Manchester / Munich?  Could that work, or do travel diaries really have to be for individuals across consecutive days, rather than just demographic 'types' which might be more or less contribute to trip summaries for each day.

3. To be consistent with trip-based model, return trips (e.g. work to home) are dropped from VISTA. Therefore, home-based trips and RRT in VISTA are also factored/weighted by 2

4. Trips that either origin or destination outside Greater Melbourne Boundary are dropped from VISTA dataset 
> - *CH: perhaps it would be better to use a buffered bounding box and origin/destination latitude and longitude, to allow for some visiting destinations outside, since we model this, I believe*

5. VISTA includes all ages (unlike TRADS, which does not include kids under 5 years old)
```{r data-prep, message=FALSE}
trips_ref <- trips_ref %>% 
  mutate(scenario = "Baseline")

trips_ref <- trips_ref %>% 
  mutate(mode = case_when(
           mode == "autoDriver" ~ "Driving Car",
           mode == "autoPassenger" ~ "Car Passenger",
           mode == "pt" ~ "Public Transport",
           mode == "walk" ~ "Walking",
           mode == "bicycle" ~ "Cycling",
           TRUE ~ "Other"),
         t.factor = if_else(t.purpose %in% c("HBW", 
                                           "HBE", 
                                           "HBA", 
                                           "HBS", 
                                           "HBR", 
                                           "HBO",
                                           "RRT"), 2,1),
         time_walk = time_walk/60,
         time_bike = time_bike/60,
         time_auto = time_auto/60,
         time_pt = time_pt/60,
         time = case_when(mode=="Cycling"~time_bike,
                          mode=="Walking"~time_walk,
                          mode=="Public Transport"~time_pt,
                          mode=="Driving Car"~time_auto,
                          mode=="Car Passenger"~time_auto),
         dist = case_when(mode=="Cycling"~t.distance_bike,
                          mode=="Walking"~t.distance_walk,
                          mode=="Public Transport"~t.distance_auto,
                          mode=="Driving Car"~t.distance_auto,
                          mode=="Car Passenger"~t.distance_auto),
         time_factored = time * t.factor,
         dist_factored = dist * t.factor,
         t.weight = 1,
         p.weight = 1,
         # convert ID to character, for consistency with VISTA person IDs
         p.ID = as.character(p.ID)) %>% 
  select(p.ID, t.purpose, time, time_factored, dist, dist_factored, mode,
         scenario, t.factor, t.weight, p.weight)

trips_scenario_vista <- trips_vista %>% 
  mutate(scenario = "VISTA")


# VISTA-version specific fields for weights and areas
# 2025-02-26 confirmed that wdtripwgt_sa3 and wdperswgt_sa3 fields exist in the VISTA trips and persons data
# the 'else' option of using LGA weights is used for test purposes with open data lacking the SA3 weights
# Note: origin and destination variables are different; the full dataset has them lower case, but also has 
# SA2 and SA3 in addition to SA4.
if ("wdtripwgt_sa3" %in% names(trips_vista) & "wetripwgt_sa3" %in% names(trips_vista) &
    "wdperswgt_sa3" %in% names(pp_vista) & "weperswgt_sa3" %in% names(pp_vista)) {
  wdtripwgt <- "wdtripwgt_sa3"
  wetripwgt <- "wetripwgt_sa3"
  wdperswgt <- "wdperswgt_sa3"
  weperswgt <- "weperswgt_sa3"
  origin_area <- "origsa4_name"
  destination_area <- "destsa4_name"
} else {
  wdtripwgt <- "wdtripwgt_LGA"
  wetripwgt <- "wetripwgt_LGA"
  wdperswgt <- "wdperswgt_LGA"
  weperswgt <- "weperswgt_LGA"
  origin_area <- "origSA4_name"
  destination_area <- "destSA4_name"
}


# exclude trips outside Greater Melbourne [CARL - please check that this works as intended:
# in particular, that the origSA4_name and destSA4_name fields are present, and that the filter 
# succeeds in picking up trips from all 9 of the specified SA4s]
trips_scenario_vista <- trips_scenario_vista %>%
  filter(!!sym(origin_area) %in% c("Melbourne - Inner", "Melbourne - Inner East",
                             "Melbourne - Inner South", "Melbourne - North East",
                             "Melbourne - North West", "Melbourne - Outer East",
                             "Melbourne - South East", "Melbourne - West",
                             "Mornington Peninsula") &
           !!sym(destination_area) %in% c("Melbourne - Inner", "Melbourne - Inner East",
                             "Melbourne - Inner South", "Melbourne - North East",
                             "Melbourne - North West", "Melbourne - Outer East",
                             "Melbourne - South East", "Melbourne - West",
                             "Mornington Peninsula"))

# add purpose to VISTA trips (see Carl's work at https://github.com/jibeproject/Melbourne-Mode-Choice/blob/main/JIBE_Melbourne_Mode_Choice.md#jibe-melbourne-mode-choice-preparation-and-analysis)

# clean trip purposes, by replacing purpose/place with 'work related' where
# one is specified and the other is missing
trips_scenario_vista <- trips_scenario_vista %>%
  mutate(origplace1 = ifelse((is.na(origplace1) | str_detect(origplace1, "Unknown")) & 
                               origpurp1 == "Work Related",
                             "Workplace", 
                             origplace1),
         destplace1 = ifelse((is.na(destplace1) | str_detect(destplace1, "Unknown")) & 
                               destpurp1 == "Work Related",
                             "Workplace", 
                             destplace1),
         origpurp1 = ifelse((is.na(origpurp1) | str_detect(origpurp1, "Unknown")) & 
                              origplace1 == "Workplace",
                            "Work Related", 
                            origpurp1),
         destpurp1 = ifelse((is.na(destpurp1) | str_detect(destpurp1, "Unknown")) & 
                              destplace1 == "Workplace",
                            "Work Related", 
                            destpurp1))
  
# assign purpose
trips_scenario_vista <- trips_scenario_vista %>%
  mutate(purpose = case_when(
    origpurp2 == "Employer's Business" | destpurp2 == "Employer's Business" ~ "business",
    origpurp1 %in% c("Unknown Purpose (at start of day)", "Not Stated") | 
      destpurp1 %in% c("NA", "Not Stated")                ~ "unknown",
    origpurp1 == "At Home" & destpurp1 == "At or Go Home" ~ "RRT",
    # trips from home are HBx, with 'x' based on destination
    origpurp1 == "At Home" ~ case_when(
      destpurp1 == "Work Related"                         ~ "HBW",
      destpurp1 == "Education"                            ~ "HBE",
      destpurp1 == "Buy Something"                        ~ "HBS",
      destpurp1 == "Recreational"                         ~ "HBR",
      destpurp1 == "Other Purpose"                        ~ "HBO",
      destpurp1 %in% c("Accompany Someone","Pick-up or Drop-off Someone") ~ "HBA",
      # classify remaining trips from home using place information if clearer than purpose
      destplace1 == "Workplace"                           ~ "HBW",
      destplace1 == "Place of Education"                  ~ "HBE",
      destplace1 == "Shops"                               ~ "HBS",
      destplace1 %in% c("Recreational Place","Natural Feature", "Social Place") ~ "HBR",
      # all other trips from home are HBO
      TRUE                                                ~ "HBO"
    ),
    # trips to home are NA
    destpurp1 == "At or Go Home"                          ~ "NA",
    origpurp1 == "Work Related" | destpurp1 == "Work Related" ~ "NHBW",
    # default for trips that are neither from home nor to home 
    TRUE                                                  ~ "NHBO"
  ))

# Inspect modes
cat(trips_scenario_vista$linkmode %>% unique() %>% sort())
## In full VISTA, the following modes were present:.groups
## "Bicycle","Jogging","Mobility Scooter","Motorcycle","Other","Public Bus","School Bus","Taxi","Train","Tram","Vehicle Driver","Vehicle Passenger","Walking"

trips_scenario_vista <- trips_scenario_vista %>% 
  left_join(pp_vista, by = "persid") %>%
  filter(!purpose%in%c("NA","business","unknown"))%>%
  mutate(
    mode = case_when(
      linkmode == "Vehicle Driver"                 ~ "Driving Car",
      linkmode == "Vehicle Passenger"              ~ "Car Passenger",
      linkmode %in% c("Public Bus", "School Bus", 
                      "Train", "Tram")             ~ "Public Transport",
      linkmode == "Walking"                        ~ "Walking",
      linkmode == "Bicycle"                        ~ "Cycling",
      # 'Other' covers Jogging, Mobility Scooter, Motorcycle, Other, and Taxi in VISTA 2012-20
      TRUE                                         ~ "Other"
    ),
    t.factor = if_else(purpose %in% c("HBW", 
                                      "HBE", 
                                      "HBA", 
                                      "HBS", 
                                      "HBR", 
                                      "HBO",
                                      "RRT"), 2*7, 1*7),
    # [SP note: alternative version of time and dist based on time/dist
    # specified in the VISTA data] [CARL - please don't change anything here at this stage:
    # relates to query below for Qin and Corin]
    time = triptime / 60,  # time converted from minutes to hours
    dist = cumdist,  # distance in km
    
    # # [SP note: will the version of VISTA have been pre-processed to include
    # # these 'bike_fast_time' etc fields? If so, uncomment (and delete above)]
    # # [CARL - this is a query for Qin and Corin; no need to do anything with this
    # # commented-out code block at the moment]
    # time = case_when(mode == "Cycling"          ~ bike_fast_time/3600,
    #                  mode == "Walking"          ~ walk_fast_time/3600,
    #                  mode == "Public Transport" ~ pt_time/3600,
    #                  mode == "Driving Car"      ~ car_congested_time/3600,
    #                  mode == "Car Passenger"    ~ car_congested_time/3600),
    # dist = case_when(mode == "Cycling"          ~ bike_fast_dist/1000,
    #                  mode == "Walking"          ~ walk_fast_dist/1000,
    #                  mode == "Public Transport" ~ pt_dist/1000,
    #                  mode == "Driving Car"      ~ car_congested_dist/1000,
    #                  mode == "Car Passenger"    ~ car_congested_dist/1000),
    
    time_factored = time * t.factor,
    dist_factored = dist * t.factor,
    
    # Lines below assume there are weekday ('wd') and weekend ('we') weights, with only 
    # the right one specified for the relevant day and the other one blank or NA].
    # [2025-02-26 - confirms this assumption holds for full VISTA 2012-20 data)
    t.weight = ifelse(is.na(.data[[wdtripwgt]]) | .data[[wdtripwgt]] == "", 
                      .data[[wetripwgt]], .data[[wdtripwgt]]),
    p.weight = ifelse(is.na(.data[[wdperswgt]]) | .data[[wdperswgt]] == "", 
                      .data[[weperswgt]], .data[[wdperswgt]]),
    # convert weights to numeric format
    t.weight = as.numeric(gsub(",", "", t.weight)),
    p.weight = as.numeric(gsub(",", "", p.weight))) %>%
  
  filter(!mode %in% c("Other"))%>%
  select(p.ID = persid, t.purpose = purpose, time, time_factored, dist, 
         dist_factored, mode, scenario, t.factor, t.weight, p.weight)

trips_scenario_vista_noWeight = trips_scenario_vista %>%
  mutate(t.weight = 1,
         p.weight = 1,
         scenario = "VISTA_noWeight")

trips <- bind_rows(trips_ref, trips_scenario_vista, trips_scenario_vista_noWeight)%>%
  mutate(mode = factor(mode, levels = c("Driving Car", 
                                        "Car Passenger", 
                                        "Public Transport", 
                                        "Walking", 
                                        "Cycling", 
                                        "Other")),
         t.purpose = case_when(
           t.purpose == "HBW" ~ "Home-based-work",
           t.purpose == "HBE" ~ "Home-based-education",
           t.purpose == "HBA" ~ "Home-based-accompanying",
           t.purpose == "HBS" ~ "Home-based-shopping",
           t.purpose == "HBR" ~ "Home-based-recreation",
           t.purpose == "HBO" ~ "Home-based-other",
           t.purpose == "NHBO" ~ "Non-home-based-other",
           t.purpose == "NHBW" ~ "Non-home-based-work",
           t.purpose == "RRT" ~ "Recreational round trip",
           TRUE ~ NA_character_),
         t.purpose = factor(t.purpose, levels = c("Home-based-work", 
                                                  "Home-based-education", 
                                                  "Home-based-accompanying", 
                                                  "Home-based-shopping", 
                                                  "Home-based-recreation", 
                                                  "Home-based-other", 
                                                  "Non-home-based-other", 
                                                  "Non-home-based-work",
                                                  "Recreational round trip")))

## 'trips' fields and data structure in sample open VISTA data
# > str(trips)
# Classes ‘data.table’ and 'data.frame':	22702179 obs. of  11 variables:
#  $ p.ID         : chr  "5" "5" "5" "5" ...
#  $ t.purpose    : Factor w/ 9 levels "Home-based-work",..: 1 1 1 1 1 1 1 1 1 1 ...
#  $ time         : num  0.162 0.162 0.162 0.162 0.178 ...
#  $ time_factored: num  0.323 0.323 0.323 0.323 0.355 ...
#  $ dist         : num  5.34 5.34 5.34 5.34 7.08 ...
#  $ dist_factored: num  10.7 10.7 10.7 10.7 14.2 ...
#  $ mode         : Factor w/ 6 levels "Driving Car",..: 1 1 1 1 1 1 4 4 4 4 ...
#  $ scenario     : chr  "Baseline" "Baseline" "Baseline" "Baseline" ...
#  $ t.factor     : num  2 2 2 2 2 2 2 2 2 2 ...
#  $ t.weight     : num  1 1 1 1 1 1 1 1 1 1 ...
#  $ p.weight     : num  1 1 1 1 1 1 1 1 1 1 ...
#  - attr(*, ".internal.selfref")=<externalptr> 

## 'trips' fields and data structure
str(trips)
```

## Key Figures
```{r message = FALSE}
pop_baseline = nrow(pp_ref)
pop_vista = sum(as.numeric(gsub(",", "", pp_vista[[wdperswgt]])), na.rm = T) +
  sum(as.numeric(gsub(",", "", pp_vista[[weperswgt]])), na.rm = T)  
pop_vista_noWeight = nrow(pp_vista)

summary = trips %>% group_by(scenario) %>% summarise(trips=sum(t.factor*t.weight, na.rm=TRUE))

summary = summary %>% 
  mutate(pop = c(pop_baseline, pop_vista, pop_vista_noWeight),
         avgTrip = trips/pop)
```

```{r vista-null-weight-check}
# check for null weights in VISTA data; there are a few -- we'll count them so we can discuss
trips$t.weight %>% is.na() %>% table()
```

```{r clean-up}
rm(trips_ref, trips_scenario_vista, trips_scenario_vista_noWeight, pp_ref, pp_vista, trips_vista)
```

```{r echo = FALSE}
# SP: everything else from here on should now run as written, and produce
# some pretty tables and figures (they were created by Qin, not me!)

summary%>%
  gt() %>%
  tab_header(
    title = "Key Figures",
  ) %>%
  cols_label(
    trips = "Total Trips over a week",
    pop = "Total population",
    avgTrip = "Weekly Trip Rates Per Person"
  ) 
```

## Trips by purpose
```{r message=FALSE}
summary = trips%>%
  group_by(scenario,t.purpose)%>%
  summarise(count=sum(t.factor*t.weight, na.rm=TRUE))%>%
  mutate(percent = count/sum(count)*100)

```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = scenario, y = percent, fill = t.purpose)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by purpose",
    y = "Proportion (%)",
    x = "Scenario",
    fill = "Purpose"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + guides(fill = guide_legend(nrow = 3)) + coord_flip() +scale_y_reverse()

```



## Trips by distance bracket
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(scenario, distance_bracket) %>%
  summarise(count = sum(t.factor*t.weight, na.rm=TRUE)) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = scenario)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_dodge(width = 1),
    vjust = -0.5,
    color = "black",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) 

```
## Trips by distance bracket by mode
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight, na.rm=TRUE), .groups = 'drop') %>%
  group_by(mode, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(summary, aes(x = scenario , y = percent, fill = distance_bracket)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket by Mode",
    y = "Proportion (%)",
    x = "Transport Mode",
    fill = "Distance (km)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    #axis.text.x = element_text(face = "bold"),
    axis.text.x = element_blank(),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free_x", nrow = 5) +
  guides(fill = guide_legend(nrow = 1)) +
  coord_flip()+
  scale_y_reverse()

```

## Trip Length Distribution by Transport Mode
```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(trips, aes(x = dist, color = scenario)) +
  geom_density() +
  labs(
    title = "Trip Length Distribution",
    x = "Distance (km)",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    # panel.grid.major = element_blank(),
    # panel.grid.minor = element_blank(),
    # axis.ticks.y = element_blank(),
    # plot.title = element_text(hjust = 0.5, face = "bold"), 
    # #axis.text.y = element_blank(),
    # #axis.text.x = element_text(face = "bold"),
    # axis.text.x = element_blank(),
    # strip.placement = "outside", 
    # strip.text = element_text(face = "bold"),
    # legend.text = element_text(face = "bold"),
    # legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free") 

```

## Overall Mode share

```{r message=FALSE}
summary <- trips %>%
  group_by(scenario,mode) %>%
  summarise(count = sum(t.factor*t.weight, na.rm=TRUE)) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8,}
ggplot(summary, aes(x = scenario, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share",
    y = "Proportion (%)",
    x = "Scenario",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) +
  coord_flip() +
  scale_y_reverse()

```

## Mode share by distance bracket

```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight, na.rm=TRUE), .groups = 'drop') %>%
  group_by(distance_bracket, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8, fig.height=12}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share by Trip Distance",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + facet_wrap(~ scenario, scales = "free_x") +guides(fill = guide_legend(nrow = 1)) 

```

## Avergae Weekly Distance Travelled Per Person
Note that 

1. Distances of all modes are routed distances resulting from MATSim. For VISTA trips, the centroids of origin mesh block and destination mesh block are used for routing. *[SP note: to be confirmed that these routed distances will be calculated for VISTA]

2. In MATSim, PT is time-based routing. So for the plots, we assume PT distances is equal to Car routed distance

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(dist=sum(dist_factored*p.weight, na.rm=TRUE))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalDist=sum(dist)) %>% 
  mutate(avgDist = case_when(scenario == "Baseline" ~ totalDist/pop_baseline,
                             scenario == "VISTA" ~ totalDist/pop_vista,
                             scenario == "VISTA_noWeight" ~ totalDist/pop_vista_noWeight))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgDist, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgDist, 2),
                y = avgDist),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly distance (km) by mode per person",
       fill = "Scenario",
       x = "", y = "Km") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```


## Average Weekly Time Travelled Per Person
Note that 

1. Travel time of all modes are routed travel time resulting from MATSim. For VISTA trips, the centroids of origin mesh block and destination mesh block are used for routing. *[SP note: to be confirmed that (1) these routed distances will be calculated for VISTA]*

2. PT travel time includes access, egress time, in-vehicle time, waiting time and transfer time.

3. Car travel time are routed on congested network

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(time=sum(time_factored*p.weight, na.rm=TRUE))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalTime=sum(time)) %>% 
  mutate(avgTime = case_when(scenario == "Baseline" ~ totalTime/pop_baseline*60,
                             scenario == "VISTA" ~ totalTime/pop_vista*60,
                             scenario == "VISTA_noWeight" ~ totalTime/pop_vista_noWeight*60))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgTime, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgTime, 1),
                y = avgTime),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly minutes by mode per person",
       fill = "Scenario",
       x = "", y = "Hour") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```
