---
title: "trip_validation_melb"
author: "Marina_Qin_Steve"
date: "2025-##-##"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(data.table)
library(gt)

rm(list=ls())
```

## Load baseline trips.csv and VISTA trips
```{r load files, message=FALSE}
trips_ref <- fread('data/trips.csv')  ## to update with Melb model outputs when available

trips_vista <- fread("data/VISTA_2012-20_LGA_V1/T_VISTA_1220_LGA_V1.csv")  ## to replace with full VISTA data

pp_vista <- fread("data/VISTA_2012-20_LGA_V1/P_VISTA_1220_LGA_V1.csv")  ## to replace with full VISTA data

pp_ref <- fread("data/pp_2021.csv")  ## to update with Melb model outputs when available

```

## Data Preparation
Note that 
1. For baseline data, home-based trips and RRT are factored/weighted by 2

2. Since VISTA is a one-day diary, all trips are factored/weighted by 7 *[SP note: to confirm that the Melbourne model output will cover a full week like the Manchester model output]*

3. To be consistent with trip-based model, return trips (e.g. work to home) are dropped from TRADS. Therefore, home-based trips and RRT in TRADS are also factored/weighted by 2

4. Trips that either origin or destination outside Greater Melbourne Boundary are dropped from TRADS dataset *[SP note: not implemented yet.  To confirm whether the VISTA data will be pre-processed to include 'OriginWithinBoundary' and 'DestinationWithinBoundary' fields like the TRADS data.  If not, an additional step can be included here to exclude trips starting or ending outside the boundary.]*

5. VISTA includes all ages (unlike TRADS, which does not include kids under 5 years old)
```{r data prep, message=FALSE}
trips_ref <- trips_ref %>% 
  mutate(scenario = "Baseline")

trips_ref <- trips_ref %>% 
  mutate(mode = case_when(
           mode == "autoDriver" ~ "Driving Car",
           mode == "autoPassenger" ~ "Car Passenger",
           mode == "pt" ~ "Public Transport",
           mode == "walk" ~ "Walking",
           mode == "bicycle" ~ "Cycling",
           TRUE ~ "Other"),
         t.factor = if_else(t.purpose %in% c("HBW", 
                                           "HBE", 
                                           "HBA", 
                                           "HBS", 
                                           "HBR", 
                                           "HBO",
                                           "RRT"), 2,1),
         time_walk = time_walk/60,
         time_bike = time_bike/60,
         time_auto = time_auto/60,
         time_pt = time_pt/60,
         time = case_when(mode=="Cycling"~time_bike,
                          mode=="Walking"~time_walk,
                          mode=="Public Transport"~time_pt,
                          mode=="Driving Car"~time_auto,
                          mode=="Car Passenger"~time_auto),
         dist = case_when(mode=="Cycling"~t.distance_bike,
                          mode=="Walking"~t.distance_walk,
                          mode=="Public Transport"~t.distance_auto,
                          mode=="Driving Car"~t.distance_auto,
                          mode=="Car Passenger"~t.distance_auto),
         time_factored = time * t.factor,
         dist_factored = dist * t.factor,
         t.weight = 1,
         p.weight = 1,
         # convert ID to character, for consistency with VISTA person IDs
         p.ID = as.character(p.ID)) %>% 
  select(p.ID, t.purpose, time, time_factored, dist, dist_factored, mode,
         scenario, t.factor, t.weight, p.weight)

trips_vista <- trips_vista %>% 
  mutate(scenario = "VISTA")

# add purpose to VISTA trips (see Carl's work at https://github.com/jibeproject/Melbourne-Mode-Choice/blob/main/JIBE_Melbourne_Mode_Choice.md#jibe-melbourne-mode-choice-preparation-and-analysis)

# clean trip purposes, by replacing purpose/place with 'work related' where
# one is specified and the other is missing
trips_vista <- trips_vista %>%
  mutate(origplace1 = ifelse((is.na(origplace1) | str_detect(origplace1, "Unknown")) & 
                               origpurp1 == "Work Related",
                             "Workplace", 
                             origplace1),
         destplace1 = ifelse((is.na(destplace1) | str_detect(destplace1, "Unknown")) & 
                               destpurp1 == "Work Related",
                             "Workplace", 
                             destplace1),
         origpurp1 = ifelse((is.na(origpurp1) | str_detect(origpurp1, "Unknown")) & 
                              origplace1 == "Workplace",
                            "Work Related", 
                            origpurp1),
         destpurp1 = ifelse((is.na(destpurp1) | str_detect(destpurp1, "Unknown")) & 
                              destplace1 == "Workplace",
                            "Work Related", 
                            destpurp1))
  
# assign purpose
trips_vista <- trips_vista %>%
  mutate(purpose = case_when(
    origpurp2 == "Employer's Business" | destpurp2 == "Employer's Business" ~ "business",
    origpurp1 %in% c("Unknown Purpose (at start of day)", "Not Stated") | 
      destpurp1 %in% c("NA", "Not Stated")                ~ "unknown",
    origpurp1 == "At Home" & destpurp1 == "At or Go Home" ~ "RRT",
    # trips from home are HBx, with 'x' based on destination
    origpurp1 == "At Home" ~ case_when(
      destpurp1 == "Work Related"                         ~ "HBW",
      destpurp1 == "Education"                            ~ "HBE",
      destpurp1 == "Buy Something"                        ~ "HBS",
      destpurp1 == "Recreational"                         ~ "HBR",
      destpurp1 == "Other Purpose"                        ~ "HBO",
      destpurp1 %in% c("Accompany Someone","Pick-up or Drop-off Someone") ~ "HBA",
      # classify remaining trips from home using place information if clearer than purpose
      destplace1 == "Workplace"                           ~ "HBW",
      destplace1 == "Place of Education"                  ~ "HBE",
      destplace1 == "Shops"                               ~ "HBS",
      destplace1 %in% c("Recreational Place","Natural Feature", "Social Place") ~ "HBR",
      # all other trips from home are HBO
      TRUE                                                ~ "HBO"
    ),
    # trips to home are NA
    destpurp1 == "At or Go Home"                          ~ "NA",
    origpurp1 == "Work Related" | destpurp1 == "Work Related" ~ "NHBW",
    # default for trips that are neither from home nor to home 
    TRUE                                                  ~ "NHBO"
  ))

# # confirm modes in full VISTA version
# modes <- trips_vista$linkmode %>% unique() %>% sort()

trips_vista <- trips_vista %>% 
  left_join(pp_vista, by = "persid") %>%
  filter(!purpose%in%c("NA","business","unknown"))%>%
  mutate(
    mode = case_when(
      linkmode == "Vehicle Driver"                 ~ "Driving Car",
      linkmode == "Vehicle Passenger"              ~ "Car Passenger",
      linkmode %in% c("Public Bus", "School Bus", 
                      "Train", "Tram")             ~ "Public Transport",
      linkmode == "Walking"                        ~ "Walking",
      linkmode == "Bicycle"                        ~ "Cycling",
      # [SP note: 'Other' (in the version of VISTA I'm using) covers Jogging,
      # Mobility Scooter, Motorcycle, Other, and Taxi - confirm in full VISTA version]
      TRUE                                         ~ "Other"
    ),
    t.factor = if_else(purpose %in% c("HBW", 
                                      "HBE", 
                                      "HBA", 
                                      "HBS", 
                                      "HBR", 
                                      "HBO",
                                      "RRT"), 2*7, 1*7),
    
    # [SP note: alternative version of time and dist based on time/dist
    # specified in the VISTA data]
    time = triptime / 60,  # time converted from minutes to hours
    dist = cumdist,  # distance in km
    
    # # [SP note: will the version of VISTA have been pre-processed to include
    # # these 'bike_fast_time' etc fields? If so, uncomment (and delete above)]
    # time = case_when(mode == "Cycling"          ~ bike_fast_time/3600,
    #                  mode == "Walking"          ~ walk_fast_time/3600,
    #                  mode == "Public Transport" ~ pt_time/3600,
    #                  mode == "Driving Car"      ~ car_congested_time/3600,
    #                  mode == "Car Passenger"    ~ car_congested_time/3600),
    # dist = case_when(mode == "Cycling"          ~ bike_fast_dist/1000,
    #                  mode == "Walking"          ~ walk_fast_dist/1000,
    #                  mode == "Public Transport" ~ pt_dist/1000,
    #                  mode == "Driving Car"      ~ car_congested_dist/1000,
    #                  mode == "Car Passenger"    ~ car_congested_dist/1000),
    
    time_factored = time * t.factor,
    dist_factored = dist * t.factor,
    
    # [SP note: check names of weight fields used in full VISTA version.  Lines
    # below assume there are weekday ('wd') and weekend ('we') weights, with only 
    # the right one specified for the relevant day and the other one blank or NA]
    t.weight = ifelse(is.na(wdtripwgt_LGA) | wdtripwgt_LGA == "", 
                      wetripwgt_LGA, wdtripwgt_LGA),
    p.weight = ifelse(is.na(wdperswgt_LGA) | wdperswgt_LGA == "", 
                      weperswgt_LGA, wdperswgt_LGA),
    # convert weights to numeric format
    t.weight = as.numeric(gsub(",", "", t.weight)),
    p.weight = as.numeric(gsub(",", "", p.weight))) %>%
  
  filter(!mode %in% c("Other"))%>%
  select(p.ID = persid, t.purpose = purpose, time, time_factored, dist, 
         dist_factored, mode, scenario, t.factor, t.weight, p.weight)

trips_vista_noWeight = trips_vista %>%
  mutate(t.weight = 1,
         p.weight = 1,
         scenario = "VISTA_noWeight")

trips <- bind_rows(trips_ref, trips_vista, trips_vista_noWeight)%>%
  mutate(mode = factor(mode, levels = c("Driving Car", 
                                        "Car Passenger", 
                                        "Public Transport", 
                                        "Walking", 
                                        "Cycling", 
                                        "Other")),
         t.purpose = case_when(
           t.purpose == "HBW" ~ "Home-based-work",
           t.purpose == "HBE" ~ "Home-based-education",
           t.purpose == "HBA" ~ "Home-based-accompanying",
           t.purpose == "HBS" ~ "Home-based-shopping",
           t.purpose == "HBR" ~ "Home-based-recreation",
           t.purpose == "HBO" ~ "Home-based-other",
           t.purpose == "NHBO" ~ "Non-home-based-other",
           t.purpose == "NHBW" ~ "Non-home-based-work",
           t.purpose == "RRT" ~ "Recreational round trip",
           TRUE ~ NA_character_),
         t.purpose = factor(t.purpose, levels = c("Home-based-work", 
                                                  "Home-based-education", 
                                                  "Home-based-accompanying", 
                                                  "Home-based-shopping", 
                                                  "Home-based-recreation", 
                                                  "Home-based-other", 
                                                  "Non-home-based-other", 
                                                  "Non-home-based-work",
                                                  "Recreational round trip")))
  
rm(trips_ref, trips_vista, trips_vista_noWeight)

```

## Key Figures
```{r message = FALSE}
pop_baseline = nrow(pp_ref)
pop_vista = sum(as.numeric(gsub(",", "", pp_vista$wdperswgt_LGA)), na.rm = T) +
  sum(as.numeric(gsub(",", "", pp_vista$weperswgt_LGA)), na.rm = T)  # [SP note: check weight fields in full VISTA as above]
pop_vista_noWeight = nrow(pp_vista)

summary = trips %>% group_by(scenario) %>% summarise(trips=sum(t.factor*t.weight))

summary = summary %>% 
  mutate(pop = c(pop_baseline, pop_vista, pop_vista_noWeight),
         avgTrip = trips/pop)

rm(pp_ref, pp_vista)
```

```{r echo = FALSE}
summary%>%
  gt() %>%
  tab_header(
    title = "Key Figures",
  ) %>%
  cols_label(
    trips = "Total Trips over a week",
    pop = "Total population",
    avgTrip = "Weekly Trip Rates Per Person"
  ) 
```

## Trips by purpose
```{r message=FALSE}
summary = trips%>%
  group_by(scenario,t.purpose)%>%
  summarise(count=sum(t.factor*t.weight))%>%
  mutate(percent = count/sum(count)*100)

```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = scenario, y = percent, fill = t.purpose)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by purpose",
    y = "Proportion (%)",
    x = "Scenario",
    fill = "Purpose"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + guides(fill = guide_legend(nrow = 3)) + coord_flip() +scale_y_reverse()

```



## Trips by distance bracket
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(scenario, distance_bracket) %>%
  summarise(count = sum(t.factor*t.weight)) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = scenario)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_dodge(width = 1),
    vjust = -0.5,
    color = "black",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) 

```
## Trips by distance bracket by mode
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight), .groups = 'drop') %>%
  group_by(mode, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(summary, aes(x = scenario , y = percent, fill = distance_bracket)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket by Mode",
    y = "Proportion (%)",
    x = "Transport Mode",
    fill = "Distance (km)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    #axis.text.x = element_text(face = "bold"),
    axis.text.x = element_blank(),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free_x", nrow = 5) +
  guides(fill = guide_legend(nrow = 1)) +
  coord_flip()+
  scale_y_reverse()

```

## Trip Length Distribution by Transport Mode
```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(trips, aes(x = dist, color = scenario)) +
  geom_density() +
  labs(
    title = "Trip Length Distribution",
    x = "Distance (km)",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    # panel.grid.major = element_blank(),
    # panel.grid.minor = element_blank(),
    # axis.ticks.y = element_blank(),
    # plot.title = element_text(hjust = 0.5, face = "bold"), 
    # #axis.text.y = element_blank(),
    # #axis.text.x = element_text(face = "bold"),
    # axis.text.x = element_blank(),
    # strip.placement = "outside", 
    # strip.text = element_text(face = "bold"),
    # legend.text = element_text(face = "bold"),
    # legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free") 

```

## Overall Mode share

```{r message=FALSE}
summary <- trips %>%
  group_by(scenario,mode) %>%
  summarise(count = sum(t.factor*t.weight)) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8,}
ggplot(summary, aes(x = scenario, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share",
    y = "Proportion (%)",
    x = "Scenario",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) +
  coord_flip() +
  scale_y_reverse()

```

## Mode share by distance bracket

```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight), .groups = 'drop') %>%
  group_by(distance_bracket, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8, fig.height=12}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share by Trip Distance",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + facet_wrap(~ scenario, scales = "free_x") +guides(fill = guide_legend(nrow = 1)) 

```

## Avergae Weekly Distance Travelled Per Person
Note that 

1. Distances of all modes are routed distances resulting from MATSim. For VISTA trips, the centroids of origin mesh block and destination mesh block are used for routing. *[SP note: to be confirmed that (1) these routed distances will be calculated for VISTA, and (2) mesh block centroids are used (I think it's correct that the locations specified in the full VISTA version are mesh block centroids; to be confirmed)]*

2. In MATSim, PT is time-based routing. So for the plots, we assume PT distances is equal to Car routed distance

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(dist=sum(dist_factored*p.weight))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalDist=sum(dist)) %>% 
  mutate(avgDist = case_when(scenario == "Baseline" ~ totalDist/pop_baseline,
                             scenario == "VISTA" ~ totalDist/pop_vista,
                             scenario == "VISTA_noWeight" ~ totalDist/pop_vista_noWeight))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgDist, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgDist, 2),
                y = avgDist),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly distance (km) by mode per person",
       fill = "Scenario",
       x = "", y = "Km") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```


## Avergae Weekly Time Travelled Per Person
Note that 

1. Travel time of all modes are routed travel time resulting from MATSim. For VISTA trips, the centroids of origin mesh block and destination mesh block are used for routing. *[SP note: to be confirmed that (1) these routed distances will be calculated for VISTA, and (2) mesh block centroids are used (I think it's correct that the locations specified in the full VISTA version are mesh block centroids; to be confirmed)]*

2. PT travel time includes access, egress time, in-vehicle time, waiting time and transfer time.

3. Car travel time are routed on congested network

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(time=sum(time_factored*p.weight))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalTime=sum(time)) %>% 
  mutate(avgTime = case_when(scenario == "Baseline" ~ totalTime/pop_baseline*60,
                             scenario == "VISTA" ~ totalTime/pop_vista*60,
                             scenario == "VISTA_noWeight" ~ totalTime/pop_vista_noWeight*60))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgTime, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgTime, 1),
                y = avgTime),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly minutes by mode per person",
       fill = "Scenario",
       x = "", y = "Hour") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```
