---
title: "trip_validation"
author: "Marina_Qin"
date: "2024-12-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(data.table)
library(gt)

rm(list=ls())
```

## Load baseline trips.csv and TRADS trips
```{r load files, message=FALSE}
trips_ref <- fread('data/trips.csv')

trips_trads <- fread("data/tripsForValidation.csv")

pp_trads <- fread("data/personForValidation.csv")

pp_ref <- fread("data/pp_2021.csv")

```

## Data Preparation
Note that 
1. For baseline data, home-based trips and RRT are factored/weighted by 2

2. Since TRADS is a one-day diary, all trips are factored/weighted by 7 

3. To be consistent with trip-based model, return trips (e.g. work to home) are dropped from TRADS. Therefore, home-based trips and RRT in TRADS are also factored/weighted by 2

4. Trips that either origin or destination outside Greater Manchester Boundary are dropped from TRADS dataset

5. TRADS does not include kids under 5 years old
```{r data prep, message=FALSE}
trips_ref <- trips_ref %>% 
  mutate(scenario = "Baseline")

trips_ref <- trips_ref %>% 
  mutate(mode = case_when(
           mode == "autoDriver" ~ "Driving Car",
           mode == "autoPassenger" ~ "Car Passenger",
           mode == "pt" ~ "Public Transport",
           mode == "walk" ~ "Walking",
           mode == "bicycle" ~ "Cycling",
           TRUE ~ "Other"),
         t.factor = if_else(t.purpose %in% c("HBW", 
                                           "HBE", 
                                           "HBA", 
                                           "HBS", 
                                           "HBR", 
                                           "HBO",
                                           "RRT"), 2,1),
         time_walk = time_walk/60,
         time_bike = time_bike/60,
         time_auto = time_auto/60,
         time_pt = time_pt/60,
         time = case_when(mode=="Cycling"~time_bike,
                          mode=="Walking"~time_walk,
                          mode=="Public Transport"~time_pt,
                          mode=="Driving Car"~time_auto,
                          mode=="Car Passenger"~time_auto),
         dist = case_when(mode=="Cycling"~t.distance_bike,
                          mode=="Walking"~t.distance_walk,
                          mode=="Public Transport"~t.distance_auto,
                          mode=="Driving Car"~t.distance_auto,
                          mode=="Car Passenger"~t.distance_auto),
         time_factored = time * t.factor,
         dist_factored = dist * t.factor,
         t.weight = 1,
         p.weight = 1)%>%
  select(p.ID,t.purpose,time,time_factored,dist,dist_factored,mode,scenario,t.factor,t.weight,p.weight)

pp_ref_over5 = pp_ref%>%filter(age>=5)%>%filter(age>=5)
trips_ref_over5 = trips_ref %>% filter(p.ID%in%pp_ref_over5$id)
trips_ref_over5 = trips_ref_over5%>%mutate(scenario = "Baseline_over5")
trips_trads <- trips_trads %>% 
  mutate(scenario = "TRADS")

trips_trads <- trips_trads %>% 
  filter(!t.purpose%in%c("NA","business","unknown"))%>%
  mutate(mode = case_when(
           t.m_main == "Car or van driver" ~ "Driving Car",
           t.m_main == "Car or van passenger" ~ "Car Passenger",
           t.m_main %in%c("Bus, minibus, coach","Metrolink", "Train") ~ "Public Transport",
           t.m_main == "Walk" ~ "Walking",
           t.m_main == "Bicycle" ~ "Cycling",
           TRUE ~ "Other"),
         t.factor = if_else(t.purpose %in% c("HBW", 
                                           "HBE", 
                                           "HBA", 
                                           "HBS", 
                                           "HBR", 
                                           "HBO",
                                           "RRT"), 2*7,1*7),
         time = case_when(mode=="Cycling"~bike_fast_time/3600,
                          mode=="Walking"~walk_fast_time/3600,
                          mode=="Public Transport"~pt_time/3600,
                          mode=="Driving Car"~car_congested_time/3600,
                          mode=="Car Passenger"~car_congested_time/3600),
         dist = case_when(mode=="Cycling"~bike_fast_dist/1000,
                          mode=="Walking"~walk_fast_dist/1000,
                          mode=="Public Transport"~pt_dist/1000,
                          mode=="Driving Car"~car_congested_dist/1000,
                          mode=="Car Passenger"~car_congested_dist/1000),
         time_factored = time * t.factor,
         dist_factored = dist * t.factor,
         t.weight = t.expansionFactor_all,
         p.weight = hh.expansionFactor)%>%
  filter(!mode%in%c("Other"))%>%
  select(p.ID,t.purpose,time,time_factored,dist,dist_factored,mode,scenario,t.factor,t.weight,p.weight)



trips_trads_noWeight = trips_trads %>%
  mutate(t.weight = 1,
         p.weight = 1,
         scenario = "TRADS_noWeight")

trips <- bind_rows(trips_ref, trips_ref_over5,trips_trads,trips_trads_noWeight)%>%
  mutate(mode = factor(mode, levels = c("Driving Car", 
                                        "Car Passenger", 
                                        "Public Transport", 
                                        "Walking", 
                                        "Cycling", 
                                        "Other")),
         t.purpose = case_when(
           t.purpose == "HBW" ~ "Home-based-work",
           t.purpose == "HBE" ~ "Home-based-education",
           t.purpose == "HBA" ~ "Home-based-accompanying",
           t.purpose == "HBS" ~ "Home-based-shopping",
           t.purpose == "HBR" ~ "Home-based-recreation",
           t.purpose == "HBO" ~ "Home-based-other",
           t.purpose == "NHBO" ~ "Non-home-based-other",
           t.purpose == "NHBW" ~ "Non-home-based-work",
           t.purpose == "RRT" ~ "Recreational round trip",
           TRUE ~ NA_character_),
         t.purpose = factor(t.purpose, levels = c("Home-based-work", 
                                                  "Home-based-education", 
                                                  "Home-based-accompanying", 
                                                  "Home-based-shopping", 
                                                  "Home-based-recreation", 
                                                  "Home-based-other", 
                                                  "Non-home-based-other", 
                                                  "Non-home-based-work",
                                                  "Recreational round trip")))
  
#trips = trips%>%filter(t.purpose!="Recreational round trip")
#trips = trips%>%filter(scenario%in%c("TRADS","Baseline_over5"))
rm(trips_ref, trips_ref_over5,trips_trads,trips_trads_noWeight)

```

## Key Figures
```{r message = FALSE}
pop_baseline = nrow(pp_ref)
pop_baseline_over5 = nrow(pp_ref%>%filter(age>=5))
pop_trads = sum(pp_trads$hh.expansionFactor)
pop_trads_noWeight = nrow(pp_trads)

summary = trips%>%group_by(scenario)%>%summarise(trips=sum(t.factor*t.weight))

summary = summary%>%
  mutate(pop= c(pop_baseline,pop_baseline_over5,pop_trads,pop_trads_noWeight),
         avgTrip= trips/pop)
rm(pp_ref,pp_ref_over5,pp_trads)
```

```{r echo = FALSE}
summary%>%
  gt() %>%
  tab_header(
    title = "Key Figures",
  ) %>%
  cols_label(
    trips = "Total Trips over a week",
    pop = "Total population",
    avgTrip = "Weekly Trip Rates Per Person"
  ) 
```

## Trips by purpose
```{r message=FALSE}
summary = trips%>%
  group_by(scenario,t.purpose)%>%
  summarise(count=sum(t.factor*t.weight))%>%
  mutate(percent = count/sum(count)*100)

#add NTS distribution for comparison
NTS <- data.frame(
  scenario = "NTS_northwest",            
  t.purpose = unique(summary$t.purpose),   
  count = 0,                   
  percent = c(18.5,6.7,14.0,16.6,19.8,8.6,9.0,2.9,3.9) 
)

summary=rbind(summary,NTS)
```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = scenario, y = percent, fill = t.purpose)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by purpose",
    y = "Proportion (%)",
    x = "Scenario",
    fill = "Purpose"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + guides(fill = guide_legend(nrow = 3)) + coord_flip() +scale_y_reverse()

```



## Trips by distance bracket
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(scenario, distance_bracket) %>%
  summarise(count = sum(t.factor*t.weight)) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=10}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = scenario)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_dodge(width = 1),
    vjust = -0.5,
    color = "black",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) 

```
## Trips by distance bracket by mode
```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight), .groups = 'drop') %>%
  group_by(mode, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```


```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(summary, aes(x = scenario , y = percent, fill = distance_bracket)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Trips by Distance Bracket by Mode",
    y = "Proportion (%)",
    x = "Transport Mode",
    fill = "Distance (km)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    #axis.text.y = element_blank(),
    #axis.text.x = element_text(face = "bold"),
    axis.text.x = element_blank(),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free_x", nrow = 5) +
  guides(fill = guide_legend(nrow = 1)) +
  coord_flip()+
  scale_y_reverse()

```

## Trip Length Distribution by Transport Mode
```{r echo=FALSE, fig.width=8, fig.height=8}
ggplot(trips, aes(x = dist, color = scenario)) +
  geom_density() +
  labs(
    title = "Trip Length Distribution",
    x = "Distance (km)",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    # panel.grid.major = element_blank(),
    # panel.grid.minor = element_blank(),
    # axis.ticks.y = element_blank(),
    # plot.title = element_text(hjust = 0.5, face = "bold"), 
    # #axis.text.y = element_blank(),
    # #axis.text.x = element_text(face = "bold"),
    # axis.text.x = element_blank(),
    # strip.placement = "outside", 
    # strip.text = element_text(face = "bold"),
    # legend.text = element_text(face = "bold"),
    # legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + 
  facet_wrap(~ mode, scales = "free") 

```

## Overall Mode share

```{r message=FALSE}
summary <- trips %>%
  group_by(scenario,mode) %>%
  summarise(count = sum(t.factor*t.weight)) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8,}
ggplot(summary, aes(x = scenario, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share by Trip Distance",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) +guides(fill = guide_legend(nrow = 1)) +
  coord_flip() +
  scale_y_reverse()

```

## Mode share by distance bracket

```{r message=FALSE}

trips$distance_bracket <- cut(
  trips$dist,
  breaks = c(0, 1, 3, 5, 10, 20, 40, Inf),
  labels = c("0-1", "1-3", "3-5", "5-10", "10-20", "20-40", "40+"),
  right = FALSE)

summary <- trips %>%
  group_by(distance_bracket, mode, scenario) %>%
  summarise(count = sum(t.factor*t.weight), .groups = 'drop') %>%
  group_by(distance_bracket, scenario) %>%
  mutate(percent = count / sum(count) * 100)

```

```{r echo=FALSE, fig.width=8, fig.height=12}
ggplot(summary, aes(x = distance_bracket, y = percent, fill = mode)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(
    aes(label = ifelse(percent > 1, paste0(round(percent, 1), "%"), "")), # Show label only if >= 1%
    position = position_fill(vjust = 0.5), 
    color = "white",
    size = 3
  ) +
  labs(
    title = "Transport Mode Share by Trip Distance",
    y = "Proportion (%)",
    x = "Distance (km)",
    fill = "Transport Mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"), 
    axis.text.y = element_blank(),
    axis.text.x = element_text(face = "bold"),
    strip.placement = "outside", 
    strip.text = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom"
  ) + facet_wrap(~ scenario, scales = "free_x") +guides(fill = guide_legend(nrow = 1)) 

```

## Avergae Weekly Distance Travelled Per Person
Note that 

1. Distances of all modes are routed distances resulting from MATSim. For TRADS trips, the centroids of origin OA and destination OA are used for routing.

2. In MATSim, PT is time-based routing. So for the plots, we assume PT distances is equal to Car routed distance

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(dist=sum(dist_factored*p.weight))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalDist=sum(dist)) %>% 
  mutate(avgDist = case_when(scenario == "Baseline" ~ totalDist/pop_baseline,
                             scenario == "Baseline_over5" ~ totalDist/pop_baseline_over5,
                             scenario == "TRADS" ~ totalDist/pop_trads,
                             scenario == "TRADS_noWeight" ~ totalDist/pop_trads_noWeight))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgDist, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgDist, 2),
                y = avgDist),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly distance (km) by mode per person",
       fill = "Scenario",
       x = "", y = "Km") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```


## Avergae Weekly Time Travelled Per Person
Note that 

1. Travel time of all modes are routed travel time resulting from MATSim. For TRADS trips, the centroids of origin OA and destination OA are used for routing. 

2. PT travel time includes access, egress time, in-vehicle time, waiting time and transfer time.

3. Car travel time are routed on congested network

```{r message=FALSE}
summary=trips%>%
  group_by(p.ID, mode,scenario)%>%
  summarise(time=sum(time_factored*p.weight))

summary=summary%>%
  group_by(mode, scenario)%>%
  summarise(totalTime=sum(time)) %>% 
  mutate(avgTime = case_when(scenario == "Baseline" ~ totalTime/pop_baseline*60,
                             scenario == "Baseline_over5" ~ totalTime/pop_baseline_over5*60,
                             scenario == "TRADS" ~ totalTime/pop_trads*60,
                             scenario == "TRADS_noWeight" ~ totalTime/pop_trads_noWeight*60))

```

```{r echo=FALSE, fig.width=8}
ggplot(summary) +
  aes(x = mode, y = avgTime, fill = scenario) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = round(avgTime, 1),
                y = avgTime),
            size = 3, #hjust = -0.1, 
            hjust = -0.2, 
            vjust = 0.2,
            position = position_dodge(1),
            inherit.aes = TRUE) +
  labs(title = "Average weekly minutes by mode per person",
       fill = "Scenario",
       x = "", y = "Hour") +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```
